# Code style guide
This chapter deals mostly with low-level coding style. Every section starts 
with a code chunk to quickly highlight the principle of that section. Depending 
on how strongly we agree/disagree with a certain practice, we use `# good / bad` 
or `# advised / discouraged` to indicate the importance of adhering to a certain
style.  
\
After the summary, the pros and cons of certain approaches are 
discussed in more depth.
```{r, include = FALSE}
knitr::opts_chunk$set(results = 'hide') # disable eval
```

## Naming 
### Naming files
```{r, eval=FALSE}
# good
reduce_matrix.R
# bad
rM_1.r
```
At the point of creating files, it is important to choose a good name. The same 
applies for the directory where it is stored. Changing the name or location of 
a file later can cause a number of problems: 

* outdated dependencies due to invalid paths.
* removed history in the version control system (e.g. not with `git mv`) if 
renaming is not done properly.
^[See [this](http://rogerdudler.github.io/git-guide/) website for a quick 
introduction to git, a popular version control system.]
* discontinuity in the history. It is hard to find renamed files.
* processes that take the file name as an input and return an output that is 
  related to this name are problematic. For example, when creating an html file 
  from an Rmd file via `knitr()`, the created object has the same name (but a 
  different suffix) as the Rmd file. If you change the name, the old html file 
  will still be there but a new html file with the new name will also be there. 
  Most likely, you need to clean up manually.

Hence, you should avoid renaming files if possible. When naming files, the 
following considerations are worth taking into account:

* Sometimes, there are restrictions in place for file names (e.g. the package 
  `testthat` used in package development requires that all files pertaining to 
  testing start with "test" and live in a particular folder).
* Do use expressive but concise names. Describe the function of the file 
  (`clean_temperature_data.R`), what it inputs or outputs 
  (`daily_to_monthly_transformation.R`), how it relates to other files (e.g. 
  `helper_sort.R` is a helper function for a sort routine) and so on.
* Do not use version numbers in the file name. Use a version control system 
  like git instead.
* Use all lower case names. One motivation for this is that it can save 
  keystrokes and improving typing efficiency. Possibly more important though is 
  that when recalling functions (given that R is case-sensitive), either 
  yourself or other R developers and users will have a harder time remembering 
  what case you used for a function name. Instead of using camelCase, separate 
  elements of the name with an underscore.
* It can be advantageous to use a combination of numbers and descriptive words 
  as a file name. This allows you to a) select files efficiently with bash or 
  even keyboard navigation in a user interface and b) can be informative if you 
  want to maintain a certain order of the files (e.g. if scripts have to be run
  in a certain) order. However, note that if you use this system, you need to 
  change the name of files if you want to insert an element at a position other 
  than the end of the sequence.
* Since some systems are not case-sensitive, do not use file names that only 
  differ in the case of the characters (e.g. `readme.md` and `README.md`).
* Choose unique file names, even across directories. This avoids confusion and
  allows you to find the file you are looking for very quickly with a search 
  function or *go to* command such as the one in RStudio.
* Use only alpha-numeric characters where possible and avoid using special 
  characters. While Linux only forbids the use of / in file names, Microsoft 
  forbids <,>,:,",/,\,|,? and * in file names. So while it may be possible to 
  use a range of special characters on some platforms, it is ill advised to do 
  so and will likely cause compatibility issues and headaches for others. The 
  only special characters you should really need to make use of are _ and -.
* Files with extremely long names might be very expressive, but that comes at 
  a cost: they are not memorable, often truncated for displaying and hence not
  really helpful anymore.
* Short and concise names are preferable but do refrain from using 
  abbreviations since that makes it hard for people to understand them if 
  they lack prior or implicit knowledge about your code. If it is absolutely 
  inevitable, provide a dictionary and a reference to it in the readme file.
* stick to a format if the content you are naming has one i.e. if your files 
  refer to specific dates or times, always use the same representation, e.g 
  YYMMDD. Other examples are names (firstname_lastname).
* Do not use synonyms, mix British or American English etc. 
* Always use the same case for file extensions. Although `xzy.r` and `xyz.R` 
  are both recognized as R files, we recommend the latter.

To sum it up, here are a couple of bad examples for the cases described 
above. 
```{r, eval = FALSE}
# version number in file name
cleaning_3.R
# special character
is_it_correct/.Rmd

# length
this_name_is_expressive_but_will_always_be_truncated_and_hence_useless.pdf
# (uncommon) abbreviations
vp8rm_fs.R
```
Do not use these pairs of examples simultaneously.
```{r, eval = FALSE}
# avoid synonyms
quick_data_check.R 
fast_data_check.R
# stick to one format
2016-02-15_tidy.RDS
15_Feb2016.csv
```
On the other hand, these file names meet the criteria outlined before.
```{r, eval = FALSE}
# lower case names
showcase_outlier.Rmd
# underscore as a separator
combine_a_and_b.R
# files in order
01-get_data.R
02-clean_data.R
03-visualize_data.R
# uppercase extension for R
for_mint.R
```

### Naming objects
```{r, eval = FALSE}
# good
reduce_matrix # for a function
reduced_matrix # for a variable
# bad
convert.characters # resembles S3 method
```
A lot of rules that were described in the above section apply for naming objects 
(in the sense of bindings within namespaces) as well. In particular we would 
like to highlight the following: 

* use verbs for functions and nouns for variables. This helps you distinguish 
  objects into categories.
* do not use the dot within object names. It gives them the appearance of 
  methods
* Avoid function names and variable names that already exist or have a 
  particular meaning. For example, `data()` is a function in base R, so it 
  might be easier to call your data `x`, `dta`, `raw` etc. Here are a couple of 
  other examples of bad R object names.
```{r}
# bad
F <- NULL
list <- "not an actual list"
mean <- "I mean he is not nice, he is quite the opposite"
sum <- function(...) summarize(...)
```
* Think about how scoping rules might affect your code and name wisely. 


## Assignment
```{r}
# good
a <- 2
# bad 
a = 2
```
Although programmers coming from other languages may not immediately see
the benefit of using the so called *assignment operator* instead of the equals
sign, the main advantage is that that an *asignment context* can be 
distinguished with ease from other contexts, for example from a function call, 
from a list creation and from a comparison of two objects.
```{r, eval = FALSE}
f(a = 1, b = TRUE)
list(a = 1:3, 
     b = sample(letters[1:4]))
a == b
```


## Quotes
```{r}
# advised
"double quotes"
# discouraged
'single quotes'
```

In R, both double quotes and single quotes are available. The advantage of 
single quotes is that they result in a slightly cleaner visual representation of
the code. 
```{r}
print('this is nice'); print("this is a bit less clean")
```
On the other hand, double quotes can be used to enclose single quotes, whereas
the reverse is not possible. This can be useful to create a string like the 
following.
```{r}
varname <- "index"
found <- sample(c("n't ", ""), size = 1)
print(paste("the variable '", varname, "' was", found, " found", sep = ""))
```
Note that in this example, single quotes behave just like any other character. 
Hence, is not necessary to *close* an open single quote. 
This allows us to also use them as contraction in `varname` without escape. 
Depending on whether or not compatibility with other programming languages is 
required, double quotes are probably also a safer option in certain cases.  
\
For these reasons, we advise the use of double quotes, so all strings are 
wrapped in the same type of quotes.

## Line indentation
Indenting code makes it more legible. For example, you can see where a 
conditional statement starts and where it ends.
```{r, include = FALSE}
condition <- TRUE
another_condition <- TRUE
```

```{r}
# good 
if (condition == TRUE){
  # quite a bit of code
  # ...
  if (another_condition == FALSE){
    # we are going even deeper...
  }
} else {
  # another block of code
}
```
Proper spacing is related to indentation and helps if a lot of consecutive lines are similar
but not exactly the same. Below, the difference of the two lines can be spotted
pretty much immediately.
```{r}
# good
big_range   <- sample(1,   size = 1,   replace = FALSE)
small_range <- sample(100, size = 100, replace = TRUE)
```
That is because some extra space was added in certain places. Without this extra
spaces, the code would look less compact.
```{r}
# bad
big_range <- sample(1, size = 1, replace = FALSE)
small_range <- sample(100, size = 100, replace = TRUE)
```
Another place to make use of indentation is when vectors are created
```{r}
# good
height <- list(Daniel = 1.98,
               Melody = 1.69,
               Peter  = 1.87)
```
There are two ways to indent: either using tabs or using spaces. We suggest two 
spaces for indentation, but in certain cases as the last example, it might be 
wise to use a more flexible indentation rule, i.e. to allow all elements of the 
list to appear exactly in line.
\
For RStudio users: By default, RStudio inserts spaces when you type tabs, so 
you can have the best of both worlds. If you prefer to press tab instead of 
spaces, just make sure that RStudio inserts the same amount of spaces as other 
people you collaborate with use. Otherwise, it will mess up the indentation. 
Regarding the more flexible indentation mentioned above, RStudio takes care of 
that too.

## Nested code
Line indentation is great, but if you hit the third level of nesting, you 
should maybe consider an approach that does not require so many sub-levels of
code. 

```{r, eval = FALSE}
# bad
if (x == TRUE){
  for (i in 4:13){
    if (z[i] == TRUE){
      # ...
    } else if (t == TRUE){
# .... 
```
In many cases, you can use functions to avoid nesting, in certain cases, the 
specific functions you want to create are *recursive functions*.^[See an example
of how to compute a [fibbonacci series](http://adv-r.had.co.nz/Function-operators.html)]
A great way to get rid of nested code related to for loops is to replace them 
with functionals. We do not want to dive into that further here, since Hadley 
Wickham's book [Advanced R](http://adv-r.had.co.nz/Functionals.html) covers the 
material quite nicely. However, let us just quickly discuss an example in the 
context of good style.
```{r}
# good
lapply(mtcars, class)
# bad
for (i in seq_len(ncol(mtcars))){
  print(class(mtcars[, i]))
}
```
Note that using a functional reduced the complexity of the code dramatically, 
which is certainly beneficial in terms of style.
Next, let us consider a functional that uses an anonymous function.^[See an 
introduction to anonymous functions in [Advanced R](http://adv-r.had.co.nz/Functional-programming.html#anonymous-functions)] One
has to take a second glance to realize that the anonymous function is wrapped 
into another function. Also, note the expression `character(1)` that 
is supplied an argument of `vapply` and does not belong to the anonymous 
function.
```{r}
x <- as.list(1:12)
vapply(x, function(r){
    if (r == 12){
      print("this is 12")
    } else {
      print("not 12")
    }
  }, character(1)) 
```

## Conditional statements
Much has been written about conditional statements. Since they are such an essential 
element of R programming and programming in general, it should be part of this 
book. In principal, R lets you write a whole *if-else-statement* in one line. 
That is four times less than using the verbose form. In the light of these, 
relationships, it might be ok to use the *one-liner* for very short statements.
```{r, include = FALSE}
fail <-    sample(c(TRUE, FALSE), 1)
success <- sample(c(TRUE, FALSE), 1)
```

```{r}
# concise
if (fail == TRUE) message("fail") else message("no fail")
# verbose
if (fail == TRUE) {
  message("fail") 
} else {
  message("no fail")
}
```
Especially for very short conditional statements, you might want to resort to 
the function `ifelse`.
```{r}
ifelse(fail == TRUE, 
       yes = "the test did fail", 
       no  = "the test did not fail")
```
`ifelse` can also take a vectorized input, which makes it even more useful to 
eliminate loops.
```{r}
x <- -3:4
# good
sqrt(ifelse(x < 0, NA, x))

# bad 
for (i in seq_along(x)){
  if (x[i] < 0){
    print(NA)
  } else {
    print(sqrt(x[i]))
  }
}
```
Note that double negation is not desirable in conditional statements, although
it technically works fine. Also, it seems quite natrual to put a *true* condition
first, or the condition that is expected to apply mostly.
```{r}
# good
if (fail == TRUE){
  message("the test failed")
}

# bad
if (fail != FALSE){
  message("the test failed")
}
```
Alternatively, you might want to define the variable `success = 1 - fail` 
instead of `fail` in order to not get confused.
```{r}
# good 
if (success == TRUE){
  message("the test succeeded")
}
# bad
if(fail != FALSE){
  message("the test did not succeede")
}
```


*If statements* can also be used in other contexts:
```{r}
paste("the test did", if(fail == FALSE) "not", "fail")
```
A very verbose alternative would be:
```{r}
if (fail == FALSE){
  "the test did not fail"
} else {
  "the test did fail"
}
```
If the duplication should be avoided, one needs to create a variable.
```{r}
if (fail == FALSE){
  msg <- "not"
} else { 
  msg <- NULL
}
paste("the test did", msg, "fail")
rm(msg)
```
When an assignment to a variable is conditional, the following is possible. 
```{r}
x <- 
  if (fail == TRUE) {
    rpois(10, lambda = 2)
  } else {
    rnorm(10)
  }
```
Boolean conditional statements can be abbreviated. The following two
*if statements* are identical. 
```{r, eval = FALSE}
if (!cond){
  # do xyz
}

if (cond == FALSE){
  # do xyz
}
```
The functions `any` and `all` are also helpful when it comes to conditional 
statements. 
```{r}
boolean <- sample(c(TRUE, FALSE), 10, replace= TRUE)
# good
if (all(boolean)){
  "all values are TRUE. That can't be true."
} else if (any(boolean)){
  "at least some values are TRUE. This is quite likely."
} else {
  "not a single TRUE value?"
}
```
Note that for conditional statements, the order matters if the 
conditions are not mutually exclusive. In other words, when all values in the 
above example are `TRUE`, then also some are. So exchanging the *if* and the 
*else if* statement will never print the message that all values are true, 
because the first condition already applies. To fix that you could be tempted 
to do the following:
```{r}
if (any(boolean) & !all(boolean)){
  "at least some values are TRUE. This is quite likely."
} else if (all(boolean)){
  "all values are TRUE. That can't be true."
} else {
  "not a single TRUE value?"
}
```
Clearly changing the ordering of the conditional statements like this creates 
redundancy and should thus be avoided.

Also, lazy evaluation can be useful in combination with conditional statements.
For example, you might want to so something within a function based on an input 
value. However, you do not know whether the value was supplied in the first 
place. Here are two ways to deal with that situation.
```{r, eval = FALSE}
# good 
if (!missing(x) && x > 10) {
  # your code
}
# bad
if (!missing(x)){
  if(x > 10 ){
   # your code 
  }
}
```

Finally, an alternative to *if statements* is the function `switch()` if the 
condition is of class character, especially if there are many conditions to 
test. Note that switch expects a match and returns an invisible `NULL` if the 
condition does not match with any option given.
```{r}
outcome <- "excelled"
switch(outcome, 
       fail  = "the test did fail", 
       just_passed = "lucky you",
       passed = "the test did not fail",
       excelled = "you must be Einstein")
```


## Argument specification
```{r}
# good
sample(100, size = 2, replace = FALSE)
# bad 
sample(100, , FALSE)
```
Whether and how many arguments you want to name depends on the context. Naming 
each argument makes the code more accessible for inexperienced R users but makes
it also less concise. Also, by definition, you do not need to specify an
argument value if you want to adhere to the default.
However, sometimes it can increase the legibility to do so, in particular when
you want to highlight that you are using a function twice and specify the 
argument differently.
```{r}
case_irrelevant <- grep("Look", "the value to look", ignore.case = TRUE)
case_relevant   <- grep("look", "the value to look", ignore.case = FALSE)
```
Regarding the order of the arguments, it is a good convention to stick to the 
default, even if you name the arguments. How long does it take you to spot the
all differences in the following two commands?
```{r}
# good
cut(3:19, breaks = 2, right = TRUE, include.lowest = FALSE)
# bad
cut(include.lowest = TRUE, breaks = 3:5, right = FALSE, x = 3:19)
```

