# Code style guide
This chapter deals mostly with low-level coding style. Every section starts 
with a code junk to quickly summarize the conclusion of the section. Depending 
on how strongly we agree/disagree of a certain practice, we use `# good / bad` 
or `# adviced / discouraged` to indicate the importance of abiding by a certain
style. 
\
After the summary, the pros and cons of certain approaches are 
discussed more in depth.

## Assignment
```{r}
# good
a <- 2
# bad 
a = 2
```
Although programmers coming from other languages may not immediately see
the benefit of using the so called *assignment operator* instead of the equal
sign, the main advantage is that that an *asignment context* can be clearly 
distinguished from other contexts, for example from a function call and from 
a comparison of two objects:
```{r, eval = FALSE}
f(a = 1, b = TRUE)
a == b
```


## Quotes
```{r, eval = FALSE}
# adviced
"double quotes"
# discouraged
'single quotes'
```

In R, both double quotes and single quotes are available. The advantage of 
single quotes is that they result in a slightly cleaner visual representation of
the code. 
```{r, eval = FALSE}
print('this is nice'); print("this is a bit less clean")
```
On the other hand, double quotes can be used to enclose single quotes, whereas
the reverse is not possible. This can be useful to create a string like the 
following:
```{r}
varname <- "index"
found <- sample(c("n't ", ""), size = 1)
print(paste("the variable '", varname, "' was", found, " found", sep = ""))
```
Note that in this example, single quotes behave just like any other character. 
Hence, is not necessary to *close* an open single quote. 
This allows us to also use them as contraction in `varname` without escape. 
Depending on whether 
or not compatibility with other programming languages is required, double 
quotes are probably also a saver option in certain cases.
\
For these reasons, we advise the use of double quotes, so all strings are 
wrapped
in the same type of quotes.